Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-25 extra/paper_original.qmd

In terms of model error, some data points show high error (see @fig-nldervisMNISTPACMAP) due to their deviation from the typical high-D data structure, acting as outliers. This is evident when comparing the images associated with these high model error points (see @fig-sampleimgerror) to others (see @fig-sampleimg).

@@

theme_set(theme_light())
imge_error_sample_within <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ factor(instance, levels = c(
    4699, 4785, 5173, 6671, 2186, 4636, 4882, 773, 2825, 6169, 6948, 5987, 1282, 7715, 622, 2989, 6912,
    4370, 1227, 4042, 2444, 836,
    1654)), ncol = 6) +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm'),
        strip.text = element_blank(),
        aspect.ratio = 1)
```

>>>>

There are certain data points that exhibit high error rates due to their deviation from the usual high-dimensional data structure, which makes them outliers. These outliers can be classified into two types: those that are outliers within the nonlinear structure and those that lie outside of it. The images associated with high model error points within the nonlinear structure display different patterns of the digit 1, as shown in @fig-sampleimgerrorwithin. However, when comparing these images to the ones found outside of the nonlinear structure, it becomes evident that the latter display different patterns of the digit 1 (see @fig-sampleimgerror). @fig-nldervisMNISTPACMAP shows some data points that have high error rates due to their outlier status.

@@

theme_set(theme_light())
imge_error_sample_within <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ factor(instance, levels = c(
    4699, 4785, 5173, 6671, 2186, 4636, 4882, 773, 2825, 6169, 6948, 5987, 1282, 7715, 622, 2989, 6912,
    4370, 1227, 4042, 2444, 836,
    1654)), ncol = 12) +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm'),
        strip.text = element_blank(),
        aspect.ratio = 1)
```

### 2025-06-25 motivationletter/motivationletter.qmd

---
author: Jayani P. Gamage
qualifications: B.Sc. (Hons) in Statistics
position: PhD student
www: jayanilakshika.netlify.app/
email: jayani.piyadigamage@monash.edu
phone: +61 4 5229 8443
#signature: sigfile.png
address:
  - Editors
  - Journal of Computational and Graphical Statistics
opening: "Dear Professor Yuguo Chen and Professor Laura M. Sangalli,"
closing: "Sincerely"
linestretch: 1.4
format: letter-pdf
---

\bigskip
Please consider our manuscript "Looking at Non-Linear Dimension Reductions as Models in the Data Space" for publication in the Journal of Computational and Graphical Statistics.

>>>>

---
author: Jayani P. Gamage
qualifications: B.Sc. (Hons) in Statistics
position: PhD student
www: jayanilakshika.netlify.app
email: jayani.piyadigamage@monash.edu
phone: +61 4 5229 8443
#signature: sigfile.png
address:
  - Editors
  - Journal of Computational and Graphical Statistics
opening: "Dear Professor Yuguo Chen and Professor Laura M. Sangalli,"
closing: "Sincerely"
linestretch: 1.4
format: letter-pdf
---

Please consider our manuscript entitled *“Looking at Non-Linear Dimension Reductions as Models in the Data Space”* for publication in the *Journal of Computational and Graphical Statistics*.

Non-linear dimension reduction (NLDR) techniques, such as tSNE and UMAP, are widely used for producing low-dimensional representations of high-dimensional data. While these methods can provide concise and powerful summaries, they also have the potential to exaggerate random patterns or create misleading structures. Different NLDR methods and different choices of hyper-parameters often produce wildly varying representations from the same data, leaving analysts with difficult decisions about which, if any, of these views are accurate.

Our manuscript introduces an approach that treats NLDR outputs as models and visualizes them directly in the original high-dimensional space using tours (movies of linear projections). This allows analysts to assess how well the NLDR model fits across all subspaces, identify mismatches, and compare the summaries provided by different methods. The approach we describe offer a new, visually interpretable way to diagnose NLDR results and support more trustworthy use of these techniques in high-dimensional data analysis.

We believe that this work is a strong fit for *JCGS*, given the journal’s focus on innovative statistical graphics, computational methodology, and tools that help analysts better understand high-dimensional data. Our approach combines statistical modeling, dynamic visualization, and interactive diagnostics, contributing to the journal’s mission of advancing the use of computation and graphics in statistical practice.

### 2025-06-25 paper.qmd

The model fitting can be adjusted using these parameters: 
- hexagon bin parameters
    - bottom left bin position $(s_1, \ s_2)$, 
    - the number of bins in the horizontal direction ($b_1$), which controls the number of bins the vertical direction ($b_2$), total number of bins ($b$), and total number of non-empty bins ($m$).
- bin density cutoff, to possibly remove low-density hexagons.

@@

#### Handling of low density bins
Bin density is computed as $n_h/n$. To make comparisons between hexagon configurations made from different parameter choices we use the number of **non-empty** bins ($m$) rather than the total number of bins.

@@

The benchmark value for removing low-density hexagons ranges between $0$ and $1$. When analyzing how these benchmark values influence model performance, it's essential to observe the change in RMSE as the benchmark value increases (@fig-param-two-curvy (c)). The RMSE shows a gradual decrease as the benchmark value goes from $1$ to $0$. Evaluating this rate of increase is important. If the increment is not considerable, the decision might lean towards retaining low-density hexagons.

<!-- I'll be honest I've no idea what this section is -->
The area of a hexagon is defined as $A = 3\sqrt{3}l^2/2$ where $l$ is the side length of the hexagon. If we know $a_1$ and $a_2$, $l$ can be computed (see appendix). The density of a hexagon grid is calculated as $\sum^{h}_{i=1}n_h/A$ and the proportion is $\sum^{h}_{i=1}n_h/Ab$. The baseline proportion is the proportion density at the smallest possible value of $a_1$. The relative proportion density is the ratio of the observed proportion density to the baseline proportion density.
```{r}
#| label: rm-lwd-bin-error

>>>>

The model fitting can be adjusted using these parameters:

- hexagon bin parameters
    - bottom left bin position $(s_1, \ s_2)$, 
    - the number of bins in the horizontal direction ($b_1$), which controls the number of bins the vertical direction ($b_2$), total number of bins ($b$), and total number of non-empty bins ($m$).
- bin density cutoff, to possibly remove low-density hexagons.

@@

#### Handling of low density bins

Standardised bin counts is computed as $w_h = n_h/n, ~~h=1, \dots m$, and $n$ is the number of observations. Density is computed as $d_h = w_h/A$ where $A$ is the area of the hexagon. As a measure of the denseness of a single grid, we can compute the average bin density, $\bar{d_h}$. This can be used to compare grids.

These quantities can be used to assess and compare the models resulting from different binwidths:

- RMSE should decrease when binwidth decreases, as the binning gets closer to observations being in their own bin. But a big drop in RMSE would indicate the lower binwidth is substantially better than the larger one.  
- Bins with no observations or a small number might be dropped from the model. This will create the wireframe not extending into sparse areas, or allowing for holes in the data to be better captured. Note that, all observations are used for the RMSE calculation, though. RMSE can be examined relative to dropping a fraction of the low density bins.
- The proportion of non-empty bins is interesting to examine across different binwidths. A good binning should have just the right amount of bins to neatly cover the shape of the data, and no more or less. 
- Lastly, an ideal distribution of the density of a grid is uniform, in the sense that if each bin captures a similar number of observations, then it has just the right number of bins to neatly cover the shape of the data. To examine this, the average bin density is compared against the range of binwidths.

<!-- 
To make comparisons between hexagon configurations made from different parameter choices we use the number of **non-empty** bins ($m$) rather than the total number of bins.

@@

The benchmark value for removing low-density hexagons ranges between $0$ and $1$. When analyzing how these benchmark values influence model performance, it's essential to observe the change in RMSE as the benchmark value increases (@fig-param-two-curvy (c)). The RMSE shows a gradual decrease as the benchmark value goes from $1$ to $0$. Evaluating this rate of increase is important. If the increment is not considerable, the decision might lean towards retaining low-density hexagons.
-->

<!-- I'll be honest I've no idea what this section is 
The area of a hexagon is defined as $A = 3\sqrt{3}l^2/2$ where $l$ is the side length of the hexagon. If we know $a_1$ and $a_2$, $l$ can be computed (see appendix). The density of a hexagon grid is calculated as $\sum^{h}_{i=1}n_h/A$ and the proportion is $\sum^{h}_{i=1}n_h/Ab$. The baseline proportion is the proportion density at the smallest possible value of $a_1$. The relative proportion density is the ratio of the observed proportion density to the baseline proportion density.-->

```{r}
#| label: rm-lwd-bin-error
